---
title: "GPHY489 Final Project"
author: "Sejal Prachand"
format: 
  html:
    embed-resources: true
    toc: true
    toc-float: true
---

## Objectives


## Dependencies and file locations

By nature, project relies heavily on the Forest Vegetation Simulator (FVS; [more information here](https://www.fs.usda.gov/managing-land/forest-management/fvs/software)). Unfortunately, this software is currently only available for Windows. It is (or was) possible to build FVS on Linux, but the Forest Service has removed build instructions for non-Windows systems as of April 2026. 

To make this project portable across systems, I have therefore included the code to re-run FVS, but it is evaluated only if the OS is Windows AND the user has FVS installed on their machine. Otherwise, FVS simulation results are read from .RDS files bundled with this project. These .RDS files were created with the code included in this document.

Similarly, I'm developing an R package that provides a more user-friendly interface for running FVS from R. This package, of course, requires an FVS installation for full functionality. Nonetheless, the package will load without warnings on non-Windows systems, and I use some system-agnostic functions in this code. Windows-only code is demarcated with the comment `##### Windows only ######` at the beginning of the code chunk.

The below (folded) chunk checks system OS. If the OS is Windows, it checks for an FVS installation in the home (`C:`) directory and messages with results. If needed, the search directory can be modified by hand (unfortunately, `utils::menu` does not work in Quarto/Rmd because code is run non-interactively). If FVS and our specific variant (FVS-IE) are found in the system, FVS code will be run. Otherwise, simulation results will be loaded from `.RDS` files.

```{r}
#| label: check-FVS
#| code-fold: true

sys_os <- tolower(Sys.info()['sysname'])

if(sys_os == 'windows'){
  ## MODIFY IF FVS IS NOT AT THIS LEVEL ##
  root <- 'C:'
  ########################################
  fvs_dir <- file.path(root, 'FVS', 'FVSSoftware')
  
  message('Looking for FVS/FVSSoftware in ', root, ' modify if necessary.')
  if(dir.exists(fvs_dir)){
    message('FVS found at ', fvs_dir, ' checking for FVSIE install...')
    if(dir.exists(file.path(fvs_dir, 'FVSbin', 'FVSie.dll'))){
      message('FVSIE found at ', file.path(fvs_dir, 'FVSbin', 'FVSie.dll'), 
              ' FVS will be run.')
    }else{
      message('FVSIE not found at ', file.path(fvs_dir, 'FVSbin', 'FVSie.dll'), 
              ' FVS will not be run.')
    }
  }else{
    message('FVS/FVSSoftware not found in ', root, '. FVS results will be loaded from .RDS files and FVS will not be run.')
  }
}else{
  message('Detected System OS is not Windows. FVS will not be run and simulation results will be loaded from .RDS files')
}
```

Now, we can load dependencies. I assume that these packages, `DBI`, and `RSQLite` are all installed on the users machine (they will be installed along with `rFVSIEtools` as dependencies for that package). Otherwise, run `install.packages()` to install any necessary packages.

```{r}
#| label: load-deps
#| collapse: true

library(here) # find files, helpful for quarto documents
library(dplyr) # data wrangling
library(ggplot2) # pretty plots
library(terra) # work with spatial data
library(sf)
library(rnaturalearth) # state outlines

# if my custom package is not installed, get it from GitHub (it's not on CRAN)
if(system.file(package = 'rFVSIEtools') == ''){
    message('Installing development package rFVSIEtools from Github')
    remotes::install_github(repo = 'sprachan/FVSIE-tools')
}

library(rFVSIEtools)
```

Now that we have dependencies loaded, we will establish some file paths that will be used across the code. These are all relative to the project directory.

```{r}
fia_path <- here('data', 'raw_data')
fvs_ready <- here('data', 'fvs_ready')
```



## Prepare FIA data

Forest Inventory and Analysis state-level databases are freely available from the [FIA Datamart](https://research.fs.usda.gov/products/dataandtools/fia-datamart). The Montana and Idaho databases (downloaded April 2026) are included as .db files in this package for full reproducibility.

First, we read in the FIA data, filtering to only forested plots that were measured after the new, standardized inventory protocol was established in 2001.

```{r}
# Data directories and files
fia_path <- here('data', 'raw_data')
fvs_ready <- list(MT = NULL, ID = NULL)
states <- c('MT', 'ID')

for(i in seq_along(states)){
  db_path <- here(fia_path, paste0('SQLite_FIADB_', states[i], '.db'))
  fvs_ready[[i]] <- fetch_cond(db_loc = db_path, 'COND_STATUS_CD == 1, CONDPROP_UNADJ == 1, INVYR > 2001, DSTRBCD1 == 0, TRTCD1 == 0') |>
    # use _ placeholder to pipe condition table into get_FIA_state()
    get_FIA_state(db_path, fia_cond_subset = _, add_identifier = TRUE)
}

# Unlist fvs_ready stuff into separate stand and tree dataframes
FVS_standInit <- bind_rows(fvs_ready$MT$FVS_StandInit,
                           fvs_ready$ID$FVS_StandInit) |>
  arrange(PID, INV_YEAR) |>
  group_by(PID) |>
  mutate(REM_CD = seq(0, n()-1),
         N_REM = max(REM_CD),
         FOREST = stringr::str_pad(FOREST, 2, pad = '0', side = 'left')) |>
  filter(VARIANT == 'IE') |>
  ungroup()

FVS_treeInit <- bind_rows(fvs_ready$MT$FVS_TreeInit,
                          fvs_ready$ID$FVS_TreeInit) |>
  left_join(FVS_standInit[c('PID', 'INV_YEAR', 'REM_CD',  'N_REM')],
            by = c('PID', 'INV_YEAR')) |>
  # Only undamaged trees that are Douglas-fir or western larch
  filter(DAMAGE1 == 0|is.na(DAMAGE1),
         SPECIES %in% c(202, 73))

rm(fvs_ready) # large list can go away
```

We only want to run FVS on stands that have been remeasured multiple times. These will be input into FVS and projected forward to the next time they were measured. For example, suppose plot X was first measured in 2003, then remeasured in 2010. Then, FVS will project growth from 2003 to 2010, allowing us to compare FVS-projected growth to FIA remeasured growth over the same time period. For simplicity, we will only project from first measurement to first *re*measurement and avoid the complexities of projecting plots from first to second *re*measurements.

```{r}
#| label: filter_stands
t0_stands <- FVS_standInit |>
  # N_REM is Number of REMeasurements; 
  #> REMeasurement CoDe is 0 for install, 1 for first remeasure, 2 for second, etc
  filter(N_REM >= 1, REM_CD == 0) |>
  dplyr::group_by(PID) |>
  # year of first remeasurement
  dplyr::mutate(REM_YEAR = ifelse(N_REM == 1, max(INV_YEAR), INV_YEAR[2])) |>
  dplyr::ungroup() |>
  as.data.frame()

# only want trees associated with selected stands
t0_trees <- FVS_treeInit |>
  inner_join(t0_stands['STAND_CN'], by = 'STAND_CN') |>
  filter(HISTORY == 1, # alive at first remeasurement only
         # large trees only
         DIAMETER >= 3) 
```

### Convert FIA locations to spatial data format

Now that we have FIA data in hand, we need to convert stand locations into a spatial dataframe. 

The documentation for FIA data indicated that the coordinates were collected in the NAD83 datum, so we will make these data into a simple features dataframe with that datum:

```{r}
stand_locs <- t0_stands |>
  select(LATITUDE, LONGITUDE, PID, OWNCD, INV_YEAR, N_REM, REM_CD) |>
  st_as_sf(coords = c('LONGITUDE', 'LATITUDE'),
           crs = st_crs('epsg:4269')) 
```

Because FIA plot locations are fuzzed for privacy, we want get climatic water deficits from buffers (polygons, not points) that contain all possible actual plot locations. Public plots are fuzzed less than private plots, so we need to account for that in our buffer creation. I'm not sure how to get `sf` objects to play nice with `SpatRaster` objects, so I'll convert stand locations to `SpatVector ` objects and use those to make buffers.

```{r}
# buffer each plot locations 0.5mi (if on public land) or 1mi (if on private land) 
pub_radius <- 0.5*1609.344 # convert mi to m
priv_radius <- 1609.344

pub_buffer <- stand_locs[stand_locs$OWNCD < 40,] |>
  vect() |>
  buffer(width = pub_radius, quadsegs = 16) #quadsegs makes these a little smoother
priv_buffer <- stand_locs[stand_locs$OWNCD >= 40,] |>
  vect() |>
  buffer(width = priv_radius, quadsegs = 16)
```

Now, we are ready to use these buffers to extract climatic water deficit for each plot location.

## Prepare climatic water deficit data

```{r}
study_region <- rnaturalearth::ne_states(country = 'united states of america') |>
  filter(iso_3166_2 %in% c('US-MT', 'US-ID')) |>
  st_transform(crs = st_crs(stand_locs))

deficit <- rast(here('data', 'raw_data', 
                     'V_1_5_annual_gridmet_historical_Deficit_1980_2019_annual_means_cropped_units_mm.tif')) |>
  terra::project(y = crs(stand_locs)) |>
  crop(ext(study_region))
```

We'll double check that everything matches up appropriately by, of course, making a map! I'll take advantage of `terra`'s built-in plotting functionality.

```{r}
plot(deficit)
lines(vect(study_region), col = 'lightgrey')
plot(pub_buffer, add = TRUE, col = 'white', border = 'white')
plot(priv_buffer, add = TRUE, col = 'white', border = 'white')
```

This looks good, so we can proceed to extracting deficit. I'm not sure how to do this in `sf`, but I do know how to do it in `terra`, so I'll be using those tools.

```{r}
pub_deficits <- extract(deficit, pub_buffer, fun = mean, touches = TRUE, bind = TRUE) |>
  as.data.frame()
priv_deficits <- extract(deficit, priv_buffer, fun = mean, touches = TRUE, bind = TRUE) |>
  as.data.frame()

stand_deficits <- bind_rows(pub_deficits, priv_deficits)
colnames(stand_deficits)[length(colnames(stand_deficits))] <- 'deficit'
```


## Run FVS

```{r}
#| label: run-FVS
#| code-fold: true

############ Windows only #################

if(sys_os == 'windows'){
  # Projection parameters
  num_years <- 11
  triple <- FALSE
  calibrate <- FALSE
  regen <- FALSE
  STDIDENT <- 'FIA_UNCALIB'
  random_seed <- 2025
  
  # Directory for .key, .tre, and .out files
  dt <- paste0('fia_uc_', format(Sys.Date(), '%b%d%Y'), '_', 
               format(Sys.time(), '%H%M'), collapse = '')
  out_dir <- file.path(fvs_runs, dt)
  dir.create(out_dir)
  
  future::plan('multisession', workers = 5)

  sim <- run_FVS_parallel(stand_info = t0_stands, 
                          tree_list = t0_trees, 
                          n_batches = 4, 
                          simple_output = TRUE, 
                          out_dir = out_dir, fvs_bin = fvs_bin, 
                          year_col = 'REM_YEAR',
                          proj_len = num_years, triple = triple,
                          calibrate = calibrate, add_regen = regen,
                          STDIDENT = STDIDENT, random_seed = random_seed)
  
  tl_fvs <- sim$tree_list |>
    group_by(PID) |>
    mutate(REM_CD = case_when(year == min(year) ~ 0, year == max(year) ~ 1, 
                              .default = NA)) |>
    rename(DIAMETER = dbh,
           YEAR = year)
  
  tl_fia <- treeInit |>
    dplyr::filter(TUID %in% tl_fvs1$TUID, REM_CD < 2) |>
    rename(YEAR = INV_YEAR)
  
  compare_growth <- full_join(tl_fvs1[c('DIAMETER', 'YEAR', 'TUID', 'REM_CD', 'PID')], 
                              tl_fia1[c('DIAMETER', 'YEAR', 'TUID', 'REM_CD', 'PID', 'HISTORY', 'DAMAGE1', 'SPECIES')], 
                              by = c('TUID', 'REM_CD', 'PID', 'YEAR'),
                              suffix = c('_FVS', '_FIA')) |>
    # only want trees that have both FIA and FVS (re)measurements/projections, respectively
    dplyr::filter(!is.na(DIAMETER_FVS), !is.na(DIAMETER_FIA)) |>
    group_by(TUID, PID, SPECIES) |>
    dplyr::summarize(growth_pd = YEAR[2]-YEAR[1],
                     status_1 = HISTORY[1],
                     status_2 = HISTORY[2],
                     dg_FVS = DIAMETER_FVS[2]-DIAMETER_FVS[1],
                     dg_FIA = round(DIAMETER_FIA[2]-DIAMETER_FIA[1], 2),
                     initial_dbh = DIAMETER_FIA[1]) |>
    filter(dg_FIA >= 0) |>
    dplyr::ungroup()
}else{
  message('Detected OS not Windows; FVS not run.')
}
```

If not on Windows, we will retrieve simulation outputs:
```{r}

```

