---
title: "Write Up"
format: html
editor: visual
---

```{r}
# Functions and Directories ----
source('../00_functions.R')

# FVS directories
fvs_dir <- file.path('C:', 'FVS', 'FVSSoftware')
fvs_bin <- file.path(fvs_dir, 'FVSbin')
fvs_rVer <- list.files(file.path(fvs_dir, 'R')) # R version used for naming dirs
rFVS_dir <- file.path(fvs_dir, 'R', fvs_rVer, 'library') # rFVS location

# Data directories and files
fia_path <- file.path('..', '..', 'data', 'fia')
fia_sql <- file.path(fia_path, 'SQLite_FIADB_MT.db')
fvs_ready <- file.path('..', '..', 'data', 'fvs_ready')
fvs_runs <- file.path('..', '..', 'data', 'fvs_runFiles')

# Packages -----
library('rFVS', lib.loc = rFVS_dir) # rFVS
library(dplyr)
library(ggplot2)
library(usmap)
library(terra)
library(sf)

# Read in remeasurement data ----
standInit <- readRDS(file.path(fvs_ready, 'FVSstandInitAll.rds'))
treeInit <- readRDS(file.path(fvs_ready, 'FVStreeInitAll.rds'))
head(standInit, 5)
head(treeInit, 5)

# Species table, for reference
s <- unique(treeInit$SPECIES)
conn <- DBI::dbConnect(RSQLite::SQLite(), fia_sql)
species_tab <- tbl(conn, 'REF_SPECIES') |>
  filter(SPCD %in% s) |>
  select(SPCD, COMMON_NAME, SPECIES_SYMBOL) |>
  collect()
DBI::dbDisconnect(conn)
```

Let's only work with plots that are meant to be used with the Inland Empire Variant (IE) and in the Northern Rockies and Bitterroot Valley ecoregion:

```{r}
standInitIE <- filter(standInit,
                      VARIANT == 'IE',
                      substr(ECOREGION, 1, 5) == 'M332B')

```

This gets us down to `{r} nrow(treeInitIE)` tree observations from `{r} nrow(standInitIE)` plot observations.

Somewhat infuriatingly, the tree table does not have measurement year/inventory year. We will have to add this as a column using information in standInitIE:

```{r}
treeInitIE <- treeInit[treeInit$STAND_CN %in% standInitIE$STAND_CN,] |>
  left_join(select(standInitIE, STAND_CN, INV_YEAR), by = join_by(STAND_CN))
```

## Exploratory Data Analysis

First, let's take a look at where these plots are in the state:

```{r}
mt_map <- us_map(regions = 'county', include = 'Montana') # state outline
stand_locs <- vect(standInitIE, geom = c('LONGITUDE', 'LATITUDE')) 
# FIA documentation tells us that they use NAD83 datum
crs(stand_locs) <- '+proj=longlat +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +no_defs +type=crs'

stand_locs <- project(stand_locs, crs(mt_map)) |> # project stands to montana
st_as_sf()
stand_map <- ggplot()+
  geom_sf(data = mt_map)+
  # don't want to plot the same plot on top of itself a bunch of times
  geom_sf(data = distinct(stand_locs, PID, .keep_all = TRUE), alpha = 0.25)
stand_map
```

Now, we want to create a column REM_CD that tells us whether this is the plot's first measure (REM_CD = 0), first remeasure (REM_CD = 1), or second remeasure (REM_CD = 2).

```{r}
standInitIE <- standInitIE |>
  arrange(PID, INV_YEAR) |>
  group_by(PID) |>
  mutate(REM_CD = case_when(INV_YEAR == min(INV_YEAR) ~ 0,
                            n() == 2 ~ 1,
                            INV_YEAR == max(INV_YEAR) ~ 2,
                            INV_YEAR != max(INV_YEAR) ~ 1,
                            .default = NA),
         N_REM = max(REM_CD)) |>
  ungroup()
```

Let's see what our breakdown of remeasured plots looks like.

```{r}
rem_plot <- standInitIE |>
  count(REM_CD) |>
  ggplot(aes(x = REM_CD, y = n))+
  geom_col()+
  geom_text(aes(label = n),
            vjust = -0.5)
rem_plot
```

So there are a good number of plots that have been remeasured once (REM_CD = 1), but very few that have been remeasured twice (REM_CD = 2). Let's see how many trees this corresponds with.

```{r}
# First, need to associate tree records with REM_CDs
treeInitIE <- select(standInitIE, STAND_CN, REM_CD) |>
  right_join(treeInitIE, by = join_by(STAND_CN))

# Now plot
tree_rem_plot <- count(treeInitIE, REM_CD) |>
  ggplot(aes(x = REM_CD, y = n))+
  geom_col()+
  geom_text(aes(label = n), vjust = -0.5)
tree_rem_plot
```

## Testing Design

To test the efficacy of FIA calibration, we will test FVS predictions from initial measure –\> first remeasure using 3 calibration techniques:

1.  No calibration
2.  Self calibration using growth data from first -\> second remeasurement.
3.  Calibration using OLS regression.

## Get uncalibrated predictions

Now, we need to access all stands that were measured twice, and access their information from the first remeasurement. This will be our set of initial stands to project.

```{r}
t0_stands <- filter(standInitIE, N_REM == 1, REM_CD == 0) |>
  select(-N_REM, -REM_CD) |>
  as.data.frame()

```

This is a total of `{r} nrow(t0_stands)` stands. Now we need to get the associated trees:

```{r}
t0_trees <- filter(treeInitIE, STAND_CN %in% t0_stands$STAND_CN) |>
  clean_FIA_treeList(standinfo = standInitIE) |>
  as.data.frame()
```

This gives us a total of `{r} nrow(t0_trees)` trees.

We'll make a directory to save keyword, tree, and output files.

```{r}
# make directory. Use date and time to help with ID later
dt <- paste0(format(Sys.Date(), '%b%d%Y'), '_', format(Sys.time(), '%H.%M'), collapse = '')
outdir <- file.path(fvs_runs, dt)
dir.create(outdir)
filenames <- rep(NA, nrow(t0_stands))

# projection parameters
num_years <- 11 # number of years into the future to project
use_tripling <- FALSE
use_calibration <- FALSE
use_regenmodel <- FALSE

# FVS will  overwrite each stand, so we need storage
uncalib_proj <- list()
uncalib_proj$treelist <- NULL
uncalib_proj$summary <- NULL

for(i in 1:nrow(t0_stands)){
  st <- t0_stands[i,] 
  tr <- t0_trees[t0_trees$STAND_CN == st$STAND_CN,]

  rFVS::fvsLoad("FVSie", fvs_bin)
  filenames[i] <- write.FVSfiles(trees=tr,
                             stand=st,
                             years_out=num_years,
                             calibrate=use_calibration,
                             triple=use_tripling,
                             add_regen=use_regenmodel,
                             outdir = outdir)
  
    # grow the stand
  fvsSetCmdLine(paste0("--keywordfile=",filenames[i],".key"))
  
  # get grown tree list
  fvs_output <- fvsInteractRun(AfterEM1="fetchTrees()",
                               SimEnd=fvsGetSummary) 
  names(fvs_output)
  
  # 2007 tree list in element 1, just to check
  tl1 <- fvs_output[[1]]$AfterEM1 |>
    dplyr::left_join(select(tr, fvs.TREE_ID, TUID, PID),
                     by = dplyr::join_by(id == fvs.TREE_ID))
  nrow(tl1)
  # 2017 tree list output in element 2. use left_join to match up with TUID
  tl2 <- fvs_output[[2]]$AfterEM1 |>
    dplyr::left_join(select(tr, fvs.TREE_ID, TUID, PID), 
                     by = join_by(id == fvs.TREE_ID))
  nrow(tl2)
  tl <- rbind(tl1, tl2)
  nrow(tl)
  uncalib_proj$treelist <- rbind(uncalib_proj$treelist, tl)
  
  # similar with plot summaries
  uncalib_proj$summary <- rbind(uncalib_proj$summary,
                                cbind(fvs_output[[length(fvs_output)]],
                                      data.frame(PID = st$PID)))
  
  rFVS::fvsLoad("FVSie", fvs_bin)
}
```

Let's check our outputs:

```{r}
names(fvs_output)

saveRDS(uncalib_proj$treelist, file = 'uncalib1.RData')
saveRDS(uncalib_proj$summary, file = 'uncalib1summary.RData')
```

There is an issue in run 2 where FVS is only projecting the trees that had diameter = 0.1...
