---
title: "Write Up"
format: html
editor: visual
---

```{r}
# Functions and Directories ----
source('../00_functions.R')

# FVS directories
fvs_dir <- file.path('C:', 'FVS', 'FVSSoftware')
fvs_bin <- file.path(fvs_dir, 'FVSbin')
fvs_rVer <- list.files(file.path(fvs_dir, 'R')) # R version used for naming dirs
rFVS_dir <- file.path(fvs_dir, 'R', fvs_rVer, 'library') # rFVS location

# Data directories and files
fia_path <- file.path('..', '..', 'data', 'fia')
fia_sql <- file.path(fia_path, 'SQLite_FIADB_MT.db')
fvs_ready <- file.path('..', '..', 'data', 'fvs_ready')
fvs_runs <- file.path('..', '..', 'data', 'fvs_runFiles')

# Packages -----
library('rFVS', lib.loc = rFVS_dir) # rFVS
library(dplyr)
library(ggplot2)
library(usmap)
library(terra)
library(sf)

# Read in remeasurement data ----
standInit <- readRDS(file.path(fvs_ready, 'FVSstandInitAll.rds'))
treeInit <- readRDS(file.path(fvs_ready, 'FVStreeInitAll.rds'))
head(standInit, 5)
head(treeInit, 5)

# Species table, for reference
s <- unique(treeInit$SPECIES)
conn <- DBI::dbConnect(RSQLite::SQLite(), fia_sql)
species_tab <- tbl(conn, 'REF_SPECIES') |>
  filter(SPCD %in% s) |>
  select(SPCD, COMMON_NAME, SPECIES_SYMBOL) |>
  collect()
DBI::dbDisconnect(conn)
```

Let's only work with plots that are meant to be used with the Inland Empire Variant (IE) and in the Northern Rockies and Bitterroot Valley ecoregion:

```{r}
standInitIE <- filter(standInit,
                      VARIANT == 'IE',
                      substr(ECOREGION, 1, 5) == 'M332B')

```

This gets us down to `{r} nrow(treeInitIE)` tree observations from `{r} nrow(standInitIE)` plot observations.

Somewhat infuriatingly, the tree table does not have measurement year/inventory year. We will have to add this as a column using information in standInitIE:

```{r}
treeInitIE <- treeInit[treeInit$STAND_CN %in% standInitIE$STAND_CN,] |>
  left_join(select(standInitIE, STAND_CN, INV_YEAR), by = join_by(STAND_CN)) |>
  left_join(species_tab, by = join_by(SPECIES == SPCD))
```

## Exploratory Data Analysis

First, let's take a look at where these plots are in the state:

```{r}
mt_map <- us_map(regions = 'county', include = 'Montana') # state outline
stand_locs <- vect(standInitIE, geom = c('LONGITUDE', 'LATITUDE')) 
# FIA documentation tells us that they use NAD83 datum
crs(stand_locs) <- '+proj=longlat +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +no_defs +type=crs'

stand_locs <- project(stand_locs, crs(mt_map)) |> # project stands to montana
st_as_sf()
stand_map <- ggplot()+
  geom_sf(data = mt_map)+
  # don't want to plot the same plot on top of itself a bunch of times
  geom_sf(data = distinct(stand_locs, PID, .keep_all = TRUE), alpha = 0.25)
stand_map
```

Now, we want to create a column REM_CD that tells us whether this is the plot's first measure (REM_CD = 0), first remeasure (REM_CD = 1), or second remeasure (REM_CD = 2).

```{r}
standInitIE <- standInitIE |>
  arrange(PID, INV_YEAR) |>
  group_by(PID) |>
  mutate(REM_CD = case_when(INV_YEAR == min(INV_YEAR) ~ 0,
                            n() == 2 ~ 1,
                            INV_YEAR == max(INV_YEAR) ~ 2,
                            INV_YEAR != max(INV_YEAR) ~ 1,
                            .default = NA),
         N_REM = max(REM_CD)) |>
  ungroup()
```

Let's see what our breakdown of remeasured plots looks like.

```{r}
rem_plot <- standInitIE |>
  count(REM_CD) |>
  ggplot(aes(x = REM_CD, y = n))+
  geom_col()+
  geom_text(aes(label = n),
            vjust = -0.5)
rem_plot
```

So there are a good number of plots that have been remeasured once (REM_CD = 1), but very few that have been remeasured twice (REM_CD = 2). Let's see how many trees this corresponds with.

```{r}
# First, need to associate tree records with REM_CDs
treeInitIE <- select(standInitIE, STAND_CN, REM_CD) |>
  right_join(treeInitIE, by = join_by(STAND_CN))

# Now plot
tree_rem_plot <- count(treeInitIE, REM_CD) |>
  ggplot(aes(x = REM_CD, y = n))+
  geom_col()+
  geom_text(aes(label = n), vjust = -0.5)
tree_rem_plot
```

Finally, let's break down remeasurements by species (how many species have been remeasured in each REM_CD):

```{r}
rem_species_plot <- count(treeInitIE, REM_CD, SPECIES_SYMBOL) |>
  ggplot(aes(x = SPECIES_SYMBOL, y = n))+
  geom_col()+
  facet_wrap(facets = vars(REM_CD), nrow = 3, scales = 'free_y')
rem_species_plot
```

It seems like even though we don't have many trees that have been remeasured twice, the most common species are pretty well represented in the 2nd remeasurement set.

## Testing Design

To test the efficacy of FIA calibration, we will test FVS predictions from initial measure –\> first remeasure using 3 calibration techniques:

1.  No calibration
2.  Self calibration using growth data from first -\> second remeasurement.
3.  Calibration using OLS regression.

## Get uncalibrated predictions

First, we have to load our FVS variant:

```{r}
rFVS::fvsLoad('FVSie', bin = fvs_bin)
```

Now, we need to access all stands that were measured twice, and access their information from the first remeasurement. This will be our set of initial stands to project.

```{r}
t0_stands <- filter(standInitIE, N_REM == 1, REM_CD == 0)

```

This is a total of `{r} nrow(t0_stands)` stands. Now we need to get the associated trees:

```{r}
t0_trees <- filter(treeInitIE, STAND_CN %in% t0_stands$STAND_CN) |>
  clean_FIA_treeList(standinfo = standInitIE)
```

This gives us a total of `{r} nrow(t0_trees)` trees.

We'll make a directory to save keyword, tree, and output files.

```{r}
# make directory. Use date and time to help with ID later
dt <- paste0(format(Sys.Date(), '%b%d%Y'), '_', format(Sys.time(), '%H.%M'), collapse = '')
outdir <- file.path(fvs_runs, dt)
dir.create(outdir)
```

Now, we will specify some keywords that will be passed to a function that formats keywords as FVS needs them:

```{r}
DATABASE <- list(DSNOUT = file.path(outdir, 'uncalib.db'),
                 SUMMARY= TRUE,
                 INVSTATS = TRUE)
```

And finally write our keyword and tree list files for FVS to know how to run. We generate a .key file and a .tre file, and the prefix for those files is returned; we save that as `filename`. Other parameters:

-   ncycles = 2 because the cycle length is 10 years and we want to project 20 years.

-   reporting_ints: the intervals (10 years, 20 years) that we want to get FVS to spit out results for

-   Turn off calibration, tripling, and regeneration (all set to FALSE)

```{r}
filenames <- rep(NA, nrow(t0_stands))
for(i in 1:2){
  st <- t0_stands[i,]
  tr <- dplyr::filter(t0_trees, STAND_CN == st$STAND_CN)
  filenames[i] <- write_FVS_files(trees = t0_trees, stand = t0_stand,
                            ncycles = 1,
                            reporting_ints = c(10, 20),
                            calibrate = FALSE, triple = FALSE, add_regen = FALSE,
                            STDIDENT = paste0('UNCALST',i),
                            DATABASE = DATABASE,
                            out_dir = outdir)
}
```
